---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

# Introduction

We are interested in estimating the price of maize grain across space, as that can help us in computing the profitability of different agricultural technologies. 

Here we explore methods to model maize prices, using data from across Ethiopia in 2014-2019. 



# Basic preperation

## Get the data

```{r}
prices <- read.csv("ETH_maizeprices.csv")
dim(prices)
head(prices)
table(prices$market)
```

The values we get are not numbers

```{r}
sapply(prices, class)
```

Let's see if we can make numbers out of the character variables

```{r}
test <- as.numeric(prices$maize_price)
table(prices$maize_price[is.na(test)])
```

We see that missing values are encoded as ".", but there are otherwise no strange things, we go ahead 

```{r}
test <- as.numeric(prices$maize_price)
table(prices$maize_price[is.na(test)])
# we see that missing values are shown as "." 
prices$maize_price <- test

test <- as.numeric(prices$longitude)
table(prices$longitude[is.na(test)])
prices$longitude <- test

test <- as.numeric(prices$latitude)
table(prices$latitude[is.na(test)])
prices$latitude <- test
```


## Coordinates

There are locations have missing coordinates

```{r}
u <- unique(prices[is.na(test), c("region", "market")])
u
```

You can use Google to find coordinates of the location, or its administrative area (Woreda)

```{r}
library(terra)
eth1 <- geodata::gadm("ETH", level=1)
eth2 <- geodata::gadm("ETH", level=2)
eth3 <- geodata::gadm("ETH", level=3)

geo <- crds(centroids(eth3[eth3$NAME_3=="Abobo"]))
prices[prices$market=="shebo- kire", c("longitude", "latitude")] <- geo 
geo <- crds(centroids(eth3[eth3$NAME_3=="Shay Bench"]))
prices[prices$market=="shewa- bench", c("longitude", "latitude")] <- geo 
prices <- prices[!is.na(prices$maize_price), ]
```


Perhaps you can fix the others? If not, let's remove them:

```{r}
prices <- na.omit(prices)
plot(eth2)
points(prices[, c("longitude", "latitude")], col="red", pch=20)
```

We can add dates bu using the year and the month names

```{r}
prices$mnt <- match(prices$month, tolower(month.name))
prices$date <- as.Date(paste0(prices$year, "-", prices$mnt, "-15"))
head(prices)
```


## Temporal patterns

Let's have a look at the price variation over time. A quick way to do that is to compute the mean value across locations, for each date.

```{r}
a <- aggregate(prices[,"maize_price", drop=FALSE], prices[,"date", drop=FALSE], mean, na.rm=TRUE)
plot(a)
lines(a)
```

You can see an annual pattern, with prices peaking in July-October, but there is also strong variation between years. We need to take these things into account when interpolating. 

One approach would be to decompose the time-series into a trend, seasonal variation, and noise. 

We first reshape the price data to the wide format, so that each market has a row, and such that we get missing values for months for which prices were not reported.

```{r, echo=FALSE}
pp <- prices[, c("market", "maize_price", "date", "longitude", "latitude")]
pp <- pp[order(pp$date), ]
pw <- reshape(pp,  timevar = "date",  idvar = c("market", "longitude", "latitude"), direction = "wide")
colnames(pw) <- gsub("maize_price.", "", colnames(pw))
pw[1:4, 1:6]

# checking that the columns in the correct order
all(as.Date(gsub("maize_price.", "", colnames(pw) )[-c(1:3)] )  |> order() == 1:69)

# we need a matrix of price values
markets <- pw[,1:3]
pm <- as.matrix(pw[,-c(1:3)])
rownames(pm) <- markets[,1]
```


Now we can have a look at the time series for a single site

```{r}
a <- ts(pm[1, ], start=c(2015,1), end=c(2018,12), frequency=12)
dc <- decompose(a)  
plot(dc) 
dc$seasonal[1:12]
```


We compute the seasonal component for all markets

```{r}
fseasonal <- function(x) {
  notna <- !is.na(x)
  if (sum(notna) < 24) return(rep(NA, 12))
  notna <- which(notna)

  start <- notna[1]
  s <- names(start)
  syr <- as.integer(substr(s, 1, 4))
  smth <-  as.integer(substr(s, 6, 7))

  end <- notna[length(notna)]
  e <- names(end)
  eyr <- as.integer(substr(e, 1, 4))
  emth <-  as.integer(substr(e, 6, 7))
  
  x <- x[start:end]
  a = ts(x, start=c(syr, smth), end=c(eyr, emth), frequency=12)
  a = zoo::na.StructTS(a)
  dc <- decompose(a)  
  dc$seasonal[1:12]
}

# for testing
#out <- matrix(nrow=nrow(pm), ncol=12)
#for (i in 1:nrow(pm)) out[i, ] = fseasonal(pm[i,])

out <- apply(pm, 1, fseasonal)
# apply transposes matrices 
dim(out)
out <- t(out)
```

Link the seasonal data back to the coordinates, and remove the missing cases

```{r}
s <- cbind(markets, out)
head(s)
```

National average seasonal trend

```{r}
na <- colMeans(out, na.rm=TRUE)
plot(na); lines(na)
```

We now have the seasonal variation. But we also need the spatial variation. We can express the price relative to the price in Addis Ababa.

```{r}
paddis <- pm[markets$market=="addis ababa", ]
# relative price for a location for each month
m <- pm / paddis
# relative price for a location
m <- rowMeans(m, na.rm=TRUE)
p <- cbind(s, spatial=m)
p <- na.omit(p)
```


## Interpolate 

First set up a raster to use. We use the extent of Ethiopia 

```{r}
eth_extent <- ext(eth1) |> floor()
r <- crop(rast(res=1/12), eth_extent)
```

Spatial price variation with TPS

```{r}
library(fields) 
xy <- as.matrix(p[,c("longitude", "latitude")])
tps <- Tps(xy, p$spatial)
sp <- interpolate(r, tps)
sp <- mask(sp, eth1)
plot(sp)
```


Or with inverse distance weighting

```{r}
xyz <- cbind(xy, p$spatial)
x <- interpIDW(r, xyz, 2)
x <- mask(x, eth1)
plot(x)
lines(eth1)
```


Or with randomforest

```{r}
library(randomForest) 
rf <- randomForest(spatial ~ longitude + latitude, data=p)
p <- interpolate(r, rf, xyNames=c("longitude", "latitude"))
p <- mask(p, eth1)
plot(p)
```


You can use co-variates 


```{r}
library(terra)
travel <- geodata::travel_time(to="city", size=5, up=TRUE) 
clm <- geodata::worldclim_country("ETH", "bio")
area <- geodata::crop_spam("maize", "area", africa=TRUE)
yield <- geodata::crop_spam("maize", "yield", africa=TRUE)
```

1) Can you use these co-variates with a RandomForest and TPS?

2) Can you interpolate the monthly prices and show the estimated price variation for one year?


